---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, fig.width = 10, fig.height = 5, fig.keep = 'all' ,fig.path = 'figures\ ', dev = 'png')

## Loading and preprocessing the data

Loading packages
```{r}
library(ggplot2)
library(ggthemes)
```

Reading downloaded file
```{r}
activity <- read.csv("activity.csv")
```

Setting date format to help get the weekdays of the dates
```{r}
activity$date <- as.POSIXct(activity$date, "%Y%m%d")
```

Getting the days of all the dates on the dataset
```{r}
day <- weekdays(activity$date)
```

Combining the dataset with the weekday of the dates
```{r}
activity <- cbind(activity, day)
```

Viewing the processed data
```{r}
summary(activity)
```

-------------------------------------------------------------------------------

## Q1: What is mean total number of steps taken per day?

Calculating total steps taken on a day
```{r}
activityTotalSteps <- with(activity, aggregate(steps, by = list(date), sum, na.rm = TRUE))
```

Changing col names
```{r}
names(activityTotalSteps) <- c("Date", "Steps")
```

Converting the data set into a data frame to be able to use ggplot2
```{r}
totalStepsdf <- data.frame(activityTotalSteps)
```

Plotting a histogram using ggplot2
```{r}
g <- ggplot(totalStepsdf, aes(x = Steps)) + 
  geom_histogram(breaks = seq(0, 25000, by = 2500), fill = "#83CAFF", col = "black") + 
  ylim(0, 30) + 
  xlab("Total Steps Taken Per Day") + 
  ylab("Frequency") + 
  ggtitle("Total Number of Steps Taken on a Day") + 
  theme_calc(base_family = "serif")

print(g)
```

# Mean of the total number of steps/day
mean(activityTotalSteps$Steps)

# Median of the total number of steps/day
median(activityTotalSteps$Steps)

-------------------------------------------------------------------------------

## Q2: What is the average daily activity pattern?

Calculating the average number of steps taken, averaged across all days by 5-min intervals.
```{r}
averageDailyActivity <- aggregate(activity$steps, by = list(activity$interval), FUN = mean, na.rm = TRUE)
```

Changing column names
```{r}
names(averageDailyActivity) <- c("Interval", "Mean")
```

Converting data set into a dataframe
```{r}
averageActivitydf <- data.frame(averageDailyActivity)
```

Plotting on ggplot2
```{r}
da <- ggplot(averageActivitydf, mapping = aes(Interval, Mean)) + 
  geom_line(col = "blue") +
  xlab("Interval") + 
  ylab("Average Number of Steps") + 
  ggtitle("Average Number of Steps Per Interval") +
  theme_calc(base_family = "serif")

print(da)
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
averageDailyActivity[which.max(averageDailyActivity$Mean), ]$Interval
```

-------------------------------------------------------------------------------

## Q3: Inputing missing values

Number of rows with NAs values
```{r}
sum(is.na(activity$steps))
```

Matching the mean of daily activity with the missing values
```{r}
inputedSteps <- averageDailyActivity$Mean[match(activity$interval, averageDailyActivity$Interval)]
```

Transforming steps in activity if they were missing values with the filled values from above.
```{r}
activityInputed <- transform(activity, steps = ifelse(is.na(activity$steps), yes = inputedSteps, no = activity$steps))
```

Forming the new dataset with the inputed missing values.
```{r}
totalActivityInputed <- aggregate(steps ~ date, activityInputed, sum)
```

Changing col names
```{r}
names(totalActivityInputed) <- c("date", "dailySteps")
```

Test if new dataset has no NA values
```{r}
sum(is.na(totalActivityInputed$dailySteps))
```

Converting the data set into a data frame to be able to use ggplot2
```{r}
totalInputedStepsdf <- data.frame(totalActivityInputed)
```

Plotting a histogram using ggplot2
```{r}
p <- ggplot(totalInputedStepsdf, aes(x = dailySteps)) + 
  geom_histogram(breaks = seq(0, 25000, by = 2500), fill = "#83CAFF", col = "black") + 
  ylim(0, 30) + 
  xlab("Total Steps Taken Per Day") + 
  ylab("Frequency") + 
  ggtitle("Total Number of Steps Taken on a Day") + 
  theme_calc(base_family = "serif")

print(p)
```

# Mean of the total number of steps taken per day
```{r}
mean(totalActivityInputed$dailySteps)
```

# Median of the total number of steps taken per day
```{r}
median(totalActivityInputed$dailySteps)
```

## Q4: Are there differences in activity patterns between weekdays and weekends?

Updating format of the dates
```{r}
activity$date <- as.Date(strptime(activity$date, format="%Y-%m-%d"))
```

Creating a function that distinguishes weekdays from weekends
```{r}
activity$dayType <- sapply(activity$date, function(x) {
  if(weekdays(x) == "Saturday" | weekdays(x) == "Sunday")
  {y <- "Weekend"}
  else {y <- "Weekday"}
  y
})
```

Creating the data set that will be plotted
```{r}
activityByDay <-  aggregate(steps ~ interval + dayType, activity, mean, na.rm = TRUE)
```

Plotting using ggplot2
```{r}
dayPlot <-  ggplot(activityByDay, aes(x = interval , y = steps, color = dayType)) + 
  geom_line() + ggtitle("Average Daily Steps by Day Type") + 
  xlab("Interval") + 
  ylab("Average Number of Steps") +
  facet_wrap(~dayType, ncol = 1, nrow=2) +
  scale_color_discrete(name = "Day Type") +
  theme_calc(base_family = "serif")

print(dayPlot) 
```


